---
title: "R for Data Science Skill Assessment"
author: "Anna Lewkowicz"
date: "2022-08-20"
output:
  html_document:
    code_folding: hide
    theme: united
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library (readr)
library(stringr)
library(ggplot2)
library(plotly)
library(dplyr)
library(magrittr)
library(scales)
library(shiny)
library(kableExtra)
library(Hmisc)
library(glue)

urlfile="https://raw.githubusercontent.com/Bioinformatics-Research-Network/skill-assessments/main/R%20for%20Data%20Science/gapminder_clean.csv"

mydata<-read_csv(url(urlfile))

names(mydata) <-str_replace_all(names(mydata), c(" " = "." ))
```
&nbsp;

This is a report of the analysis of gapminder_clean.csv data.

# Analysis of data from year 1962

```{r mydata, echo=FALSE}
mydata_1962 <- mydata %>%
  filter(Year == 1962)
```
&nbsp;


```{r mydata_1962, fig.align="center", warning=FALSE, message=FALSE}
plot_1962 <- ggplot(mydata_1962, aes(x=`CO2.emissions.(metric.tons.per.capita)` ,y=gdpPercap)) + 
  geom_point(color = "firebrick") + 
  ggtitle("The correlation between CO2 emissions and GDP per capita in year 1962") + 
  labs(y= "GDP per capita", x = expression("CO2 emissions (metric tons per capita)")) +
  theme(axis.title.x = element_text(vjust = 0, size = 15),
        axis.title.y = element_text(vjust = 2, size = 15),
        axis.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5))

print(plot_1962)
```
&nbsp;

Pearson's correlation of CO2 emissions and GDP per capita in 1962 was calculated and the resuls are presented below.
&nbsp;


```{r mydata_1962_correlation}

cor.test(mydata_1962$`CO2.emissions.(metric.tons.per.capita)`, mydata_1962$gdpPercap,)

```
&nbsp;

The correlation of 'CO2 emissions (metric tons per capita)' and gdpPercap equals 0.9260817. 

The associated p-value equals < 2.2e-16.

# In what year is the correlation between CO2 emissions and GDP per capita the strongest?
&nbsp;

```{r mydata_correlation}

all_years <- unique(mydata$Year)

year_cor_co2_gdp <- setNames(data.frame(matrix(ncol = 2, nrow = 0)), 
                             c("Year", "Correlation"))

for (year in all_years) {
  subset <- mydata %>%
    filter(Year == year)
  correlation <- cor(subset$`CO2.emissions.(metric.tons.per.capita)`, 
                     subset$gdpPercap, use = "complete.obs")
  year_cor_co2_gdp[nrow(year_cor_co2_gdp) + 1,] <- c(year, correlation)
}

year_cor_co2_gdp[order(year_cor_co2_gdp$Correlation, decreasing = TRUE),] %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```
&nbsp;

The correlation between 'CO2 emissions (metric tons per capita)' and gdpPercap is the strongest in the year 1967.
&nbsp;


```{r mydata_1967, fig.align="center"}

mydata_1967 <- mydata %>%
  filter(Year == 1967)

plot_1967 <- ggplot(mydata_1967, aes(x=`CO2.emissions.(metric.tons.per.capita)` ,y=gdpPercap)) + 
  geom_point(aes(size = pop, colour = continent)) + 
  ggtitle("The correlation between CO2 emissions and GDP per capita in year 1967") + 
  labs(y= "GDP per capita", x = "CO2 emissions (metric tons per capita)") +
  theme(axis.title.x = element_text(vjust = 0, size = 15),
        axis.title.y = element_text(vjust = 2, size = 15),
        axis.text = element_text(size = 10)) 
div(ggplotly(plot_1967), align = "center")
```

# What is the relationship between continent and energy use?
&nbsp;

One-way ANOVA was chosen to determine what is the relationship between continent and energy use. It was chosen because the examined data had one categorical independent variable (continent) that had multiple levels (Asia, Europe, Africa, Americas, Oceania) and one quantitative dependent variable (energy use). 

Results are presented below.

```{r energy_use_anova}
one_way_anova_1967 <- aov(`Energy.use.(kg.of.oil.equivalent.per.capita)` ~ continent, data = mydata_1967)

summary(one_way_anova_1967)
```
&nbsp;

The p-value is low (p < 0.001), it appears that depending on the continent, there is a difference in energy use.


# Is there is a significant difference between Europe and Asia with respect to imports of goods and services in the years after 1990?
&nbsp;

Because two groups (Europe and Asia) were being compared, t-test was chosen to assess whether there is a significant difference between Europe and Asia with respect to imports of goods and services in the years after 1990.

Results are presented below.

```{r imports_anova}
mydata_1990 <- mydata %>%
  filter(Year > 1990) %>%
  filter(continent == "Europe" | continent == "Asia")

ttest_1990 <- t.test(`Imports.of.goods.and.services.(%.of.GDP)` ~ continent, data = mydata_1990)

ttest_1990
```
&nbsp;

Because of high (>0.05) p-value it is concluded that there is no significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990.


# Which country (or countries) has the highest population density across all years?
&nbsp;

```{r highest_pop_dens}
pop_dens_avg <- setNames(data.frame(matrix(ncol = 2, nrow = 0)), 
                         c("Country", "Average.population.density"))

all_countries <- unique(mydata$Country.Name)
  
for (country in all_countries) {
  subset <- mydata %>%
    filter(Country.Name == country)
  average <- mean(as.numeric(subset$`Population.density.(people.per.sq..km.of.land.area)`), na.rm=TRUE)
  pop_dens_avg[nrow(pop_dens_avg) + 1,] <- c(country, average)
}

pop_dens_avg$Average.population.density <- as.numeric(as.character(pop_dens_avg$Average.population.density))

pop_dens_avg <- pop_dens_avg[order(pop_dens_avg$Average.population.density, decreasing = TRUE),]  

head(pop_dens_avg) %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```
&nbsp;

Macao region in China has the highest 'Population density (people per sq. km of land area)' across all years. It is equal to 14732.035.

# Which country (or countries) has shown the greatest increase in life expectancy at birth since 1962?
&nbsp;

```{r head_tail}
first_year <- head(all_years, n=1) 

last_year <- tail(all_years, n=1) 

print(glue("First measurment was taken in {first_year} and last one in {last_year}."))

```



```{r life_exp}

mydata_2007 <- mydata %>%
  filter(Year == 2007)

exp_increase <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), 
                         c("Country", "Life.exp.increase.numerical", 
                           "Life.exp.increase.percentage"))

for (country in all_countries) {
  subset_1962 <- mydata_1962 %>%
    filter(Country.Name == country)
  subset_2007 <- mydata_2007 %>%
    filter(Country.Name == country)
  increase_num <- subset_2007$`Life.expectancy.at.birth,.total.(years)` - subset_1962$`Life.expectancy.at.birth,.total.(years)`
  increase_perc <- round(subset_2007$`Life.expectancy.at.birth,.total.(years)` / subset_1962$`Life.expectancy.at.birth,.total.(years)`*100, digits=1)
  if (length(increase_num)==0) {
    increase_num <- NA
    increase_perc <- NA
  }
  exp_increase[nrow(exp_increase) + 1,] <- c(country, increase_num, increase_perc)
}

exp_increase$Life.exp.increase.numerical <- as.numeric(as.character(exp_increase$Life.exp.increase.numerical))

exp_increase$Life.exp.increase.percentage <- as.numeric(as.character(exp_increase$Life.exp.increase.percentage))
```
&nbsp;

The table below shows table ordered in descending order by column containing numerical life expectancy increase. 
&nbsp;

```{r life_exp_numerical}
head(exp_increase[order(exp_increase$Life.exp.increase.numerical, decreasing = TRUE),]) %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```
&nbsp;

The table below shows table ordered in descending order by column containing percentage life expectancy increase. 
&nbsp;

```{r life_exp_percentage}
head(exp_increase[order(exp_increase$Life.exp.increase.percentage, decreasing = TRUE),]) %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```


&nbsp;

In the Maldives life expectancy has grown by 37 years, what is a growth of 196%. In Bhutan life expectancy has grown by 33 years, what is over 200%. 